Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-24 WSSSPE1_report/WSSSPE1_summary_report.tex
small change
Updating the developing and supporting software section
Turning the notes into somewhat structured paragraphs with common
themes, once complete I will start organizing them into subsections
and adding more references to the papers that were discussed.

@@

\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted
territories. Highest risk. Overlap with the utility software as it must serve
as a facilitator to run existing software. Long requirement-development-test
cycles. Wide community support but tends to be handled by specialists rather
then end users. Promotes collaborations, one of the key ingredient of a
sustainable process.

>>>>

The panel on developing and supporting software examined the challenges
around scientific software development/support mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several stated that documentation is not just for users, and paying attention
to API documentation, tutorials for building and deploying software, along with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

There was discussion that backward compatibility is not always desirable, and it
can be very costly. This must be balanced with the aims of a given project, how
many other projects depend on and use the code when backwards incompatible
changes are to be made. There are many examples in the wider open source
software world of strategies for dealing with this, and again best practices
can go a long way to mitigating issues around backwards compatibility. Many
projects live with sub-optimal code for a while, and make allow backwards
compatibility to be broken at agreed upon development points, such as a major
release for a software library.

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the papers submitted discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as lists where developers do not participate in
mailing lists as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek
and Turk's paper on scaling code in the human dimension discuss how development
lists that are welcoming to people actually have many more people contributing.

Discussion around recruiting and/or retaining personnel in this area is hard,
with one of the major reasons being no long-term career paths (especially when
compared with industry). How should software development be balanced with
research? It is apparent that things are beginning to improve, such as
incentives for software development in the form of altmetrics, tenure committee
consideration, NSF Products vs Publications is was noted it can be very
difficult to measure where people are using small bits of your code.

There were 13 articles about different experiences in this area, but little
about GUI testing, performance, scalability, or agile development practices.
There were several unique perspectives about issues such as managing API
changes, using the same best practices for software as data, and going beyond
simply ``slapping an OSI-approved license on code''.

The question of what ``sustainable'' means in the context of software was
raised. The resources at http://oss-watch.ac.uk/resources/ssmm discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Questions such as whether the supplier will still be around in five
years time, will it still care for its customers needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

What is the overarching goal of sustainability? Is it reproducible science,
persistence, quality, something else? How should success be measured in this
context? Is there some metric that can be used to determine when you have
achieved sustainable software, or is this an ongoing process with no clear
endpoint. For truly sustainable software there should be no endpoint, as the
software products continue to be used and useful beyond the single institution,
grant, and developer/development team. Sustainability must be addressed
throughout the project life cycle.

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware''.

The question of whether funders, such as the NSF, can mandate software plans in
much the same way as they do data management plans. Daniel Katz responded that
software is supposed to be described as part of the NSF data management plan,
and that in their definition data includes software. A comment from Twitter
(@biomickwatson) raised the issue that this requires reviewers and funders who
understand the answers that are given in these plans. Daniel Katz responded
that in programs focused on software or data this can be done effectively, but
conceded that in more general programs this is indeed a problem.

The papers submitted to this panel, and several others, include lots of general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. The idea received some resistance,
stating difficulties in acting as a central common repository with so many
compelling alternatives, and as an orphanage. The point for centralization of
communication/point of contact might be reasonable, with the statement that
``vested with authority'' is perhaps too strong, but ``providing tools if
needed'' might be more appropriate. The wider open source community is moving
to a model of distributed version control (Git, Mercurial, Bazaar, etc), rather
than more centralized, as discussed in the second keynote.

Some of the questions raised after the panel discussion.

Rather than teaching developers about domain science, or domain scientists about
software development practices, why don't we teach both communities to
collaborate more effectively? Can this be done without teaching each side a
little of the other to enable communication, with a response that it really is
not two binary communities, but a spectrum with useful roles in the middle.
Someone else asked if a developer can be effective without being part of the
domain community, with responses that this really depends on the specific
problem---translators and T-shaped people are important. Why aren't academic
communities taught how to evaluate cross-disciplinary work well?

Is there a role for the growing science of the team science field in this? There
is overlap between the communities, with support for virtual organizations,
tool development, etc. How can we make time in an already crowded schedule to
introduce these topics to students? Should they be introduced through lab
classes as in ``real sciences''?

@@

\subsection{Infrastructure Software}
Infrastructure oriented software developed to better utilize new
infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted
territories. Highest risk since the projects are experimental. Overlap with the
utility software as it must serve as a facilitator to run existing software.
Long requirement-development-test cycles. Wide community support but tends to
be handled by specialists rather then end users. Promotes collaborations, one
of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures.
